Reliable feature point matching

### 2013-08-12 text/main/ch1_introduction.tex
Fixed references

\addcontentsline{toc}{chapter}{Introduction}
\chapter*{Introduction}
Matching feature points is one of the problems in computer vision that 
if solved well will provide a useful tool for improving methods on a 
host of interesting problems. If we can match feature points reliably we 
can recognize objects, make panoramas, recognize duplicate photos, guide 
robots and airplanes and a host of other interesting applications 
\footnote{For a few examples of some of these applications, check out: 
\cite{se2002global} (robot guidance), \cite{brown2007automatic} 
(panorama construction), \cite{lowe2004sift} (has parts on object 
recognition), \cite{chu2010consumer} (duplicate detection)}.  For all 
these problems better feature point matching will most likely yield 
better solutions.

@@

The thesis is organized as follows: Chapter~\ref{C:background} provides 
background and motivation for the development of \emph{MM} and 
\emph{MMC}. The chapter serves two purposes.  For those that might be 
reading this introduction wondering what a feature point really is and 
how it is matched, it provides a gentle introduction to the why's and 
how's of feature point matching. At the same time the chapter is meant 
as an introduction to related work which is hopefully useful for those 
who might use this thesis as a starting point for finding other research 
on the same subject. Chapter~\ref{C:algorithms} will introduce in detail 
the algorithms proposed in this thesis. The reference algorithms used to 
benchmark \emph{MM} and \emph{MMC} are also introduced in detail.  
Chapter~\ref{C:experiments} details the experiments done to evaluate the 
proposed algorithms both in terms of how they were conducted as well as 
their results. Chapter~\ref{C:applications} provides two examples of how 
the matching algorithms can improve on panorama stitching and near 
duplicate matching.

For the remainder of the thesis I will make use of the academic 
'\emph{we}' to underline the fact this thesis had not been possible had 
it not been for the guidance of Stefan Winkler and Sabine S\"usstrunk as 
well as the support of Advanced Digital Science Center in Singapore that 
hosted me as an intern for the duration of my thesis work.

>>>>

\addcontentsline{toc}{chapter}{Introduction}
\chapter*{Introduction}
%Matching feature points is one of the problems in computer vision that 
%if solved well will provide a useful tool for improving methods on a 
%host of interesting problems. If we can match feature points reliably 
%we can recognize objects, make panoramas, recognize duplicate photos, 
%guide robots and airplanes and a host of other interesting applications 
%\footnote{For a few examples of some of these applications, check out: 
%\cite{se2002global} (robot guidance), \cite{brown2007automatic} 
%(panorama construction), \cite{lowe2004sift} (has parts on object 
%recognition), \cite{chu2010consumer} (duplicate detection)}.  For all 
%these problems better feature point matching will most likely yield 
%better solutions.

@@

The thesis is organized as follows: Chapter~\ref{C:Background} provides 
background and motivation for the development of \emph{MM} and 
\emph{MMC}. The chapter serves two purposes.  For those that might be 
reading this introduction wondering what a feature point really is and 
how it is matched, it provides a gentle introduction to the why's and 
how's of feature point matching. At the same time the chapter is meant 
as an introduction to related work which is hopefully useful for those 
who might use this thesis as a starting point for finding other research 
on the same subject. Chapter~\ref{C:Algorithms} will introduce in detail 
the algorithms proposed in this thesis. The reference algorithms used to 
benchmark \emph{MM} and \emph{MMC} are also introduced in detail.  
Chapter~\ref{C:Experiments} details the experiments done to evaluate the 
proposed algorithms both in terms of how they were conducted as well as 
their results. Chapter~\ref{C:Applications} provides two examples of how 
the matching algorithms can improve on panorama stitching and near 
duplicate matching.

For the remainder of the thesis I will make use of the academic 
'\emph{we}' to underline the fact this thesis had not been possible had 
it not been for the guidance of Stefan Winkler and Sabine S\"usstrunk as 
well as the support of Advanced Digital Science Center in Singapore that 
hosted me as an intern during my thesis work.

### 2013-08-12 text/main/ch3_matching_methods.tex
Fixed references

\chapter{Algorithms}
\label{C:algorithms}

@@

In the \emph{acquisition stage} we extract local feature points and 
obtain their descriptors from the set of images. The algorithm is not 
designed to make use of any feature extractor in particular, but for the
experiments (Chapter~\ref{S:Experiments}) the SIFT feature extractor and
descriptor has been used. In the \emph{matching} stage these features 
are matched using $k$-nearest neighbors.  The distance function $d(f_i, 
f_j)$ between features is calculated according to the feature 
descriptor. For SIFT we use the cosine similarity between the two 
descriptor vectors, while other descriptors might use other 
measures\footnote{BRIEF for example uses the hamming measure to speed up 
the computation}.  For any given feature $f_i$ the two closest neighbors 
are returned, and we calculate the ratio between them as proposed in 
\cite{lowe2004sift} (In Part~\ref{ss:ratio} this calculation and other 
details of \emph{Ratio} will be discussed).  Any correspondence with a 
ratio above the threshold supplied will be discarded. Finally in the 
filter stage we check that matches are from different images and discard 
all matches that are not symmetric. That is; if a feature, $f_i$, is 
best matched to another feature, $f_j$, we would want for $f_j$ to be 
best matched to $f_i$ too.
This isn't always the case, for example if $f_i$ doesn't have a true 
correspondance in the other image, its closest neighbor $f_j$ might 
likely be better matched with a different feature point.

@@

To cluster the feature points, we construct a graph with the feature 
points as nodes and their similarity score as normalized edge weights.  
Given a set of $n$ features $F = {f_1, \ldots, f_n}$ and a similarity 
function $s$ we can define the matching function $M : (f_i, f_j) 
\rightarrow \mathbb{R}$ as follows:
\begin{equation*}
    M(a,b) = \begin{cases} similarity(f_i, f_j) & \mbox{if } f_i \neq 
    f_j \\ 0 & \mbox{otherwise}
	\end{cases}
\end{equation*}
The similarity function is based on the distance measure $d(f_i,f_j)$ 
(cosine similarity in the case of SIFT): $similarity(f_i, f_j) = 1 - 
d(f_i,f_j) / c$ where $c$ is a normalization constant normalizing the 
distance values between $0$ and $1$.
We obtain $A$, the normalized adjecency matrix of the fully connected 
similarity graph of feature points as follows:
\begin{equation*}
    A_{i,j} = \frac{M(f_i, f_j)}{\sum\limits_{f_i,f_j \in F} M(f_i, f_j) 
    / \left\vert F \right\vert ^ {2}}
\end{equation*}

@@

\begin{figure}[t]
    \centering
	\includegraphics[width=\textwidth]{images/MMC_graph}
    \label{fig:graph}
	\caption{The partitioned feature graph. Each vertex represents a 
		feature point; lines indicate high similarity between points. A 
		partition is a connected group with the same color. The border 
		color of each node indicates which image it belongs to.  Zooming 
	into a subsection of the graph (right part), the various cluster 
sizes can be seen, ranging from hundreds of feature points to only two 
or three.}
	\label{fig:graph}
\end{figure}

@@

While the Louvain clustering algorithm does not require any parameters 
in itself, it tends towards clustering all feature points together in 
the same partition if the graph is very connected.  To ensure that the 
graph is well clustered, the adjacency matrix is pruned setting all 
edges below a certain threshold to $0$. In Algorithm~\ref{alg-mmc} this 
threshold is provided as the variable $\alpha$.  From empirical 
analysis, retaining the top 2.5\% of edges with the highest similarity 
seems to work well in practice. This requires estimating $\alpha$ 
parameter after obtaining the adjecency matrix.  Figure~\ref{fig:graph} 
shows the result of clustering the feature points as a graph.

@@

The matching algorithm for \emph{MMC} detailed in 
Algorithm~\ref{alg-getmatches} finds matches within all partitions with 
more than two elements using the \emph{MM} approach (\emph{MMGet}).  
However, as can be seen in the example in Figure~\ref{fig:graph}, many 
of the partitions contain only two feature points from different images 
linked by one edge. In this case, we compare the similarity of the these 
two feature points with their second best match and remove matches where 
this ratio lies above a certain threshold, like in the \emph{Ratio} 
algorithm (\emph{RatioGet}).  For example in the case of 
Figure~\ref{fig:pitts_partition}, we have feature points from a building 
in one image grouped together with points from a shirt pattern in 
another.  The nearest neighbor method would have returned wrong matches, 
but since we match the partition with \emph{MM}, points in the building 
end up matching other points in the building, and no matches are 
returned.

>>>>

\chapter{Algorithms}
\label{C:Algorithms}

@@

In the \emph{acquisition stage} we extract local feature points and 
obtain their descriptors from the set of images. The algorithm is not 
designed to make use of any feature extractor in particular, but for the
experiments (Chapter~\ref{C:Experiments}) the SIFT feature extractor and
descriptor has been used. In the \emph{matching} stage these features 
are matched using $k$-nearest neighbors.  The distance function $d(f_i, 
f_j)$ between features is calculated according to the feature 
descriptor. For SIFT we use the cosine similarity between the two 
descriptor vectors, while other descriptors might use other 
measures\footnote{BRIEF for example uses the hamming measure to speed up 
the computation}.  For any given feature $f_i$ the two closest neighbors 
are returned, and we calculate the ratio between them as proposed in 
\cite{lowe2004sift}\footnote{In Part~\ref{ss:ratio} this calculation and 
other details of \emph{Ratio} will be discussed}.  Any correspondence 
with a ratio above the threshold supplied will be discarded. Finally in 
the filter stage we check that matches are from different images and 
discard all matches that are not symmetric. That is; if a feature, 
$f_i$, is best matched to another feature, $f_j$, we would want for 
$f_j$ to be best matched to $f_i$ too.
This isn't always the case, for example if $f_i$ doesn't have a true 
correspondance in the other image, its closest neighbor $f_j$ might 
likely be better matched with a different feature point.

@@

To cluster the feature points, we construct a graph with the feature 
points as nodes and their similarity score as normalized edge weights.  
Given a set of $n$ features $F = {f_1, \ldots, f_n}$ and a similarity 
function $s$ we can define the matching function $M : (f_i, f_j) 
\rightarrow \mathbb{R}$ as follows:
\begin{equation*}
    M(a,b) = \begin{cases} s(f_i, f_j) & \mbox{if } f_i \neq f_j \\ 0 & 
    \mbox{otherwise}
	\end{cases}
\end{equation*}
The similarity function $s$ is based on the distance measure 
$d(f_i,f_j)$ (cosine similarity in the case of SIFT): $s(f_i, f_j) = 1 - 
d(f_i,f_j) / c$ where $c$ is a normalization constant normalizing the 
distance values between $0$ and $1$.
We obtain $A$, the normalized adjecency matrix of the fully connected 
similarity graph of feature points as follows:
\begin{equation*}
    A_{i,j} = \frac{M(f_i, f_j)}{\sum\limits_{f_i,f_j \in F} M(f_i, f_j) 
    / \left\vert F \right\vert ^ {2}}
\end{equation*}

@@

\begin{figure}[t]
    \centering
	\includegraphics[width=\textwidth]{images/MMC_graph}
    \caption{The partitioned feature graph. Each vertex represents a 
        feature point; lines indicate high similarity between points. A 
        partition is a connected group with the same color. The border 
        color of each node indicates which image it belongs to.  Zooming 
    into a subsection of the graph (right part), the various cluster 
sizes can be seen, ranging from hundreds of feature points to only two 
or three.}
    \label{fig:graph-view}
\end{figure}

@@

While the Louvain clustering algorithm does not require any parameters 
in itself, it tends towards clustering all feature points together in 
the same partition if the graph is very connected.  To ensure that the 
graph is well clustered, the adjacency matrix is pruned setting all 
edges below a certain threshold to $0$. In Algorithm~\ref{alg-mmc} this 
threshold is provided as the variable $\alpha$.  From empirical 
analysis, retaining the top 2.5\% of edges with the highest similarity 
seems to work well in practice. This requires estimating $\alpha$ 
parameter after obtaining the adjecency matrix.  
Figure~\ref{fig:graph-view} shows the result of clustering the feature 
points as a graph.

@@

The matching algorithm for \emph{MMC} detailed in 
Algorithm~\ref{alg-getmatches} finds matches within all partitions with 
more than two elements using the \emph{MM} approach (\emph{MMGet}).  
However, as can be seen in the example in Figure~\ref{fig:graph-view}, 
many of the partitions contain only two feature points from different 
images linked by one edge. In this case, we compare the similarity of 
the these two feature points with their second best match and remove 
matches where this ratio lies above a certain threshold, like in the 
\emph{Ratio} algorithm (\emph{RatioGet}).  For example in the case of 
Figure~\ref{fig:pitts_partition}, we have feature points from a building 
in one image grouped together with points from a shirt pattern in 
another.  The nearest neighbor method would have returned wrong matches, 
but since we match the partition with \emph{MM}, points in the building 
end up matching other points in the building, and no matches are 
returned.

### 2013-08-12 text/main/ch5_experiments.tex
Fixed references

\chapter{Experiments and Results}
\label{S:Experiments}
%
To reliably measure the accuracy of a matching method on real images, we 
either need a set of image pairs tied by a homography, or we have to 
manually count the number of inliers. The latter becomes prohibitive for 
large numbers of (non-trivial) images.

@@

The comparison was done using the \emph{Murals} dataset 
(Figure~\ref{fig:murals}), the \emph{Graf} set (Figure~\ref{fig:Graf}) 
from \cite{mikolajczyk2005performance}, and two 
images\footnote{100\_1942.jpg and 100\_1941.jpg} from the Gallagher 
dataset \cite{gallagher2008} (Figure~\ref{fig:comparemirror}).

Test sets were generated from the image pairs by cropping square patches of
$250\!\times\!250$ pixels with a random vertical and horizontal offset.  
Given a source image pair, we produce 100 pairs of patches, which might or might not overlap.  
This ensures that patch pairs with no overlap still retain a general similarity to each 
other, while patches that do overlap often only share a small 
part of their area. Figure~\ref{fig:fairey} shows an example of 
possible pairs of test image patches produced from a source image pair.  
Producing $n$ such pairs allows us to test not just how well the 
matching algorithm performs on a variety of overlaps but also how many 
false positives we get on similar images that do not overlap at all. In 
practice the amount of overlap between pairs in a test set will depend 
on the overlap and viewpoint change in the source image pair.  To give a 
rough idea, Table~\ref{table:overlap} shows the overlap of patch pairs 
created from images 1 and 3 of the \emph{Graf} image set from 
\cite{mikolajczyk2005performance}.

@@

The results are presented as \emph{precision} versus \emph{recall} 
similar to \cite{ke2004pca} and \cite{mikolajczyk2005performance}.  
Recall corresponds to the number of correct matches over the total 
amount of possible correct matches in between the two images:
\begin{equation*}
	recall = \frac{\# ~ Correct ~ Matches}{\# ~ Total ~ Possible ~ 
	Matches}
\end{equation*}
This measure tells us how many thorough we are in finding all possible 
matches that could be made between the two images. To decide whether a 
match is correct, we use the following method:
Given a potential match between two pixels $p_1$ and $p_2$, $m = 
\left(p_1, p_2\right)$, and a homography $H$ relating the two images 
$I_1$ and $I_2$, we can calculate if $m$ is an inlier by checking if the 
two points satisfy the following criteria:
\begin{equation*}
\left\vert H p_1 - p_2 \right\vert + \left\vert H^{-1}p_2 - p_1 \right\vert < d_{\max}
\end{equation*}
That is, the distance between $p_1$ translated to $I_2$ and $p_2$ 
\emph{plus} the distance between $p_2$ translated to $I_1$ should be 
less than a certain threshold (we use $d_{\max}=5$ pixels here). To 
count the number of possible matches we go over all possible 
combinations of feature points between the two images, measure their 
distance and count how many lie within the threshold.
Precision on the other hand is measured by the amount of correct matches
over the total amount of matches returned:
\begin{equation*}
	precision = \frac{\# ~ Correct ~ Matches}{\# ~ Correct ~ Matches ~ + 
	~ \# ~ False ~ Matches}
\end{equation*}
This measures how accurate the set of returned matches from the 
algorithm is. An algorithm that returns only a few matches out of many, 
but most of them are correct will have a high precision and a low 
recall, where as an algorithm that returns almost all possible matches, 
both true and false will have a low precision but a high recall. The 
ideal matching algorithm will be both precise but also return as many 
correct matches as possible. However for many applications we only need 
enough correct matches to align to the images, in which case we care 
more about precision than recall.

@@

@@

In order to investigate whether a geometric algorithm can be improved by
pruning the initial matching set, we tested a hybrid algorithm 
\emph{Spectral-MMC} where matches are pruned by \emph{MMC} before being 
applied to \emph{Spectral}. The results are shown in 
Figure~\ref{fig:result_spectral-mmc} and show that \emph{Spectral-MMC} 
not only outperforms \emph{Spectral} but also improves on \emph{MMC} at 
similar recall rates, illustrating how a geometric algorithm can be 
complemented by a reliable non-geometric pruning of matches.

@@

>>>>

\chapter{Experiments and Results}
\label{C:Experiments}

In order to demonstrate that the algorithms we propose are in fact 
superior to similar algorithms in the field, we will have to compare 
them. In this chapter we will first describe exactly how we intent to 
test the algorithms and then discuss the results of these comparisons.

\section{Experiments}
For our ideal matching algorithm we would like it to return not only 
correct matches but all possible correct matches when images overlap and 
no matches at all when they don't.  We would like it to work flawlessly 
not just one a few images but on all possible images, and we would 
expect it to run fast even on when faced with tens of thousands of 
feature points.

These requirements fall in to three different categories. We want to 
verify that the algorithm is \emph{accurate}, \emph{general} and 
\emph{fast}.

\subsection{Measuring Accuracy}

To measure the accuracy of an algorithm we are interested in two values:
Precision and Recall. These measures were introduced in \cite{ke2004pca}
and later used in \cite{mikolajczyk2005performance}. Recall corresponds 
to the number of correct matches over the total amount of possible 
correct matches in between the two images:
\begin{equation*}
	recall = \frac{\# ~ Correct ~ Matches}{\# ~ Total ~ Possible ~ 
	Matches}
\end{equation*}
This measure tells us how thorough we are in finding all possible 
matches that could be made between the two images. To decide whether a 
match is correct, we use the following method:
Given a potential match between two pixels $p_1$ and $p_2$, $m = 
\left(p_1, p_2\right)$, and a homography $H$ relating the two images 
$I_1$ and $I_2$, we can calculate if $m$ is an inlier by checking if the 
two points satisfy the following criteria:
\begin{equation*}
\left\vert H p_1 - p_2 \right\vert + \left\vert H^{-1}p_2 - p_1 \right\vert < d_{\max}
\end{equation*}
That is, the distance between $p_1$ translated to $I_2$ and $p_2$ 
\emph{plus} the distance between $p_2$ translated to $I_1$ should be 
less than a certain threshold (we use $d_{\max}=5$ pixels here). To 
count the number of possible matches we go over all possible 
combinations of feature points between the two images, measure their 
distance and count how many lie within the threshold.
Precision on the other hand is measured by the amount of correct matches
over the total amount of matches returned:
\begin{equation*}
	precision = \frac{\# ~ Correct ~ Matches}{\# ~ Correct ~ Matches ~ + 
	~ \# ~ False ~ Matches}
\end{equation*}
This measures how accurate the set of returned matches from the 
algorithm is. An algorithm that returns only a few matches out of many, 
but most of them are correct will have a high precision and a low 
recall, where as an algorithm that returns almost all possible matches, 
both true and false will have a low precision but a high recall. The 
ideal matching algorithm will be both precise but also return as many 
correct matches as possible. However for many applications we only need 
enough correct matches to align to the images, in which case we care 
more about precision than recall.

\subsection{Measuring Generality}

To convincingly show that the algorithms work for general cases we need 
to demonstrate good recall and precision across a wide range of images.  
However if we are just given to images and a set of matches, there is no
way to tell which matches are correct and which aren't without going 
through them by hand and count them. For any but the simplest cases this
becomes prohibitively time consuming.

A way to circumvent this problem is to restrict the images we test on to 
image pairs that are linked by a homography. Two images linked by a 
homography are tied by a change in perspective such that we given a 
point in one image and the homography we can calculate the corresponding 
position of the point in the other image. Not all image pairs can be 
linked with a homography, something that is intuitive if we image two 
pictures of a playing child. We can't predict the positions of the hands
of the child in the second image with pixel precision just by knowing 
the position of the hands in the first image. However for images where 
all points are situated on a plane or where all the points are far away,
a homography is usually sufficent to describe the pixel transformations 
from one image to the next.

@@

To increase the generality of the image pairs to cases that partially 
overlap or have no overlap at all, we generated test sets from the image 
pairs by cropping square patches of $250\!\times\!250$ pixels with a 
random vertical and horizontal offset.  Given a source image pair, we 
produce 100 pairs of patches that might or might not overlap.  This 
ensures that patch pairs with no overlap still retain a general 
similarity to each other, while patches that do overlap often only share 
a small part of their area.  Figure~\ref{fig:fairey} shows an example of 
possible pairs of test image patches produced from a source image pair.  
Producing $n$ such pairs allows us to test not just how well the 
matching algorithm performs on a variety of overlaps but also how many 
false positives we get on similar images that do not overlap at all. In 
practice the amount of overlap between pairs in a test set will depend 
on the overlap and viewpoint change in the source image pair.  To give a 
rough idea, Table~\ref{table:overlap} shows the overlap of patch pairs 
created from images 1 and 3 of the \emph{Graf} image set from 
\cite{mikolajczyk2005performance}.

In addition to the \emph{Murals} data set we created a similar test set
based on the the \emph{Graf} set (Figure~\ref{fig:Graf}) showing 
different amounts of viewpoint change. A more typical real life test 
case was created as well using two images\footnote{100\_1942.jpg and 
100\_1941.jpg} (\emph{Pitts}) from the Gallagher dataset 
\cite{gallagher2008} (Figure~\ref{fig:comparemirror}). These two images 
are typical holiday shots and feature occlusion as well as a slight 
change in perspective.

@@

\subsection{Measuring Speed}

To gain an intuition about the typical running time of an algorithm we 
need to understand how the speed of the algorithm varies with different 
input lengths (i.e.\ pictures of different sizes in our case) as well as 
a general idea of how long it takes the algorithm to run given inputs of
a typical case. The first is usually denoted the computational 
complexity of the algorithm, while the second is a measure of speed.

Once we know how long it will take the algorithm to run for inputs of a 
certain size we can use the complexity to extrapolate how much longer it
will take to run for inputs of e.g.\ double or half that size. For 
example a matching algorithm with a complexity of $O(n^2)$ that takes 
four seconds to find matches for $100$ feature points will take most 
likely take $16$ seconds to find matches for $200$ feature points. If 
the algorithm had a complexity of $O(n)$ it would more likely take $8$ 
seconds to run on $200$ feature points.

For this reason we will have to look at both the measuring speed as well
as the computational complexity of the algorithms.

@@

Using the \emph{Murals} dataset as well as test sets created from the 
\emph{Graf} and \emph{Pitts} images we tested \emph{MM} and \emph{MMC} 
against \emph{Ratio}, \emph{Spectral} and \emph{Isomatch}.

\subsection{Accuracy and Generality}

@@

In order to investigate whether a geometric algorithm can be improved by
pruning the initial matching set, we tested a hybrid algorithm 
\emph{Spectral-MMC} where matches are selected by \emph{MMC} before 
being filtered by \emph{Spectral}. The results are shown in 
Figure~\ref{fig:result_spectral-mmc} and show that \emph{Spectral-MMC} 
not only outperforms \emph{Spectral} but also improves on \emph{MMC} at 
similar recall rates, illustrating how a geometric algorithm can be 
complemented by a reliable non-geometric pruning of matches.

@@

\subsection{Computational Complexity and Speed}
